Combinatorial Proofs and Decomposition Theorems for First-order Logic

### 2021-01-24 main.tex
tweaked intro

@@

% As a general rule, do not put math, special symbols or citations
% in the abstract
\begin{abstract}
  In this paper we uncover a close relation between combinatorial and
  syntactic proofs for first-order logic. Whereas syntactic proofs are
  formalized in some deductive proof system based on inference rules,
  a combinatorial proof is a ``syntax-free'' presentation of a proof
  that is independent from any set of inference rules. We show that
  the two proof representations are related via a deep inference
  decomposition theorem that establishes a new kind of normal form
  of syntactic proofs. As a consequence, we obtain (i) a simple proof
  of the soundness and completeness of first-order combinatorial
  proofs, and (ii) a full completeness theorem: every combinatorial
  proof is the image of a syntactic proof.
\end{abstract}

@@

First-order predicate logic is one of the cornerstones of modern
logic. Since its formalisation by Frege~\cite{frege:79} it has seen a
growing usage in many fields of mathematics and computer science. Since
the development of proof theory by Hilbert~\cite{hilbert:22},
\emph{proofs} became first-class citizens as mathematical objects that
could be studied on their own. Since Gentzen's \emph{sequent
calculus}~\cite{gentzen:35:I,gentzen:35:II}, many other proof systems
have been developed that allow the implementation of efficient proof
search, for example \emph{analytic tableaux}~\cite{smullyan:68} or
\emph{resolution}~\cite{robinson:65}. Despite the immense progress
that has been made in proof theory in general and in the area of
automated and interactive theorem provers in
particular\cite{DSAV:15lean,paulson:isabelle}\todo{find
references}\juihsuan{done}, we still have
no satisfactory notion of proof identity for first-order logic. In
this respect, proof theory is quite different from any other
mathematical field. For example in group theory, two groups are
\emph{the same} iff they are isomorphic; in topology, two spaces are
\emph{the same} iff they are homeomorphic; etc. In proof theory, we
have no such notion telling us when two proofs are \emph{the same},
even though Hilbert was considering this problem as possible 24th
problem for his famous lecture~\cite{hilbert:00} in
1900~\cite{thiele:03}, before proof theory existed as a mathematical
field.

@@

This is where \emph{combinatorial proofs} come in. They have been
introduced by Hughes~\cite{hughes:pws} for classical propositional
logic as ``syntax-free'' notion of proof, and as a possible solution
to Hilbert's 24th problem~\cite{hughes:invar} (see
also~\cite{str:hilbert:24}). The basic idea is to abstract away from
the syntax of the inference rules used in the proof and consider the
proof as a combinatorial object, more precisely as a special kind of
homomorphism between two graphs obeying certain properties.
\lutz{an example here?}

It has been shown by several authors how syntactic proofs in various
proof systems can be translated to propositional combinatorial proofs:
for sequent proofs in~\cite{hughes:invar}, for deep inference proofs
in~\cite{str:fscd17}, for Frege systems in~\cite{str:RR-9048}, and for
tableaux systems and resolution in~\cite{acc:str:18}. This allows to
define a natural notion of proof identity for propositional logic: two
proofs are \emph{the same}, if they are mapped to the same
combinatorial proof.

Recently, Acclavio and Stra\ss burger extended this notion to relevant
logics~\cite{acc:str:relevant} and to modal
logics~\cite{acc:str:modal}; and Heijlties, Hughes and Stra\ss burger
have provided combinatorial proofs for intuitionistic propositional
logic~\cite{HHS:lics19}.

In this paper we would like to push forward the idea that
combinatorial proofs can also for first order logic be used as a
notion of proof identity. \emph{First-order combinatorial proofs} have
been introduced by Hughes in~\cite{hughes:fopws}. But even though
Hughes shows that the conclusion of every first-order combinatorial
proof is a valid formula, his proof is not really satisfactory, as (i)
it is long and cumbersome, and (ii) it does not allow to read back a
syntactic proof based inference rules. In fact, there is the
fundamental problem that not all combinatorial proofs can be obtained
as translations of sequent calculus proofs.

In this paper we solve this issue by moving to a deep inference
system. More precisely, we introduce a new proof system, called
$\FOKS$, for first-order logic, that (i) reflects every combinatorial
proof, i.e., there is a surjective mapping from proofs in $\FOKS$ to
combinatorial proofs, that (ii) allows to provide simpler proofs of
soundness and completeness of combinatorial proofs, and (iii) admits
new decomposition theorems establishing a precise correspondence
between certain syntactic inference rules and certain combinatorial
notions.

@@

\input{pics.tex}
The fograph of drinker formula $\drinkerformula=\veedrinkerformula$:
%
\begin{center}\vspace{2ex}\begin{tikzpicture}
\drinkerbase
\end{tikzpicture}\vspace{1ex}\end{center}
%
A combinatorial proof of drinker formula $\drinkerformula=\veedrinkerformula$:
%
\begin{center}\vspace{2ex}\begin{tikzpicture}
\drinkerfibcoloured
\end{tikzpicture}\vspace{1ex}\end{center}
%
Condensed combinatorial proof of drinker formula(s):
%
\begin{center}\vspace{3ex}\begin{math}
\begin{tikzpicture}
\drinkerInline
\end{tikzpicture}
\hspace{10ex}
\begin{tikzpicture}
\veedrinkerInline
\end{tikzpicture}\end{math}\end{center}

>>>>

\input{pics.tex}

@@

% As a general rule, do not put math, special symbols or citations
% in the abstract
\begin{abstract}
  We uncover a close relationship between combinatorial and
  syntactic proofs for first-order logic. Whereas syntactic proofs are
  formalized in some deductive proof system based on inference rules,
  a combinatorial proof is a ``syntax-free'' presentation of a proof
  that is independent from any set of inference rules. We show that
  the two proof representations are related via a deep inference
  decomposition theorem that establishes a new kind of normal form
  of syntactic proofs. This yields (a) a simple proof
  of the soundness and completeness of first-order combinatorial
  proofs, and (b) a full completeness theorem: every combinatorial
  proof is the image of a syntactic proof.
\end{abstract}

@@

First-order predicate logic is a cornerstone of modern
logic. Since its formalisation by Frege~\cite{frege:79} it has seen a
growing usage in many fields of mathematics and computer science. Upon
the development of proof theory by Hilbert~\cite{hilbert:22},
\emph{proofs} became first-class citizens as mathematical objects that
could be studied on their own. Since Gentzen's \emph{sequent
calculus}~\cite{gentzen:35:I,gentzen:35:II}, many other proof systems
have been developed that allow the implementation of efficient proof
search, for example \emph{analytic tableaux}~\cite{smullyan:68} or
\emph{resolution}~\cite{robinson:65}. Despite the immense progress
made in proof theory in general and in the area of
automated and interactive theorem provers in
particular, we still have
no satisfactory notion of proof identity for first-order logic. In
this respect, proof theory is quite different from any other
mathematical field. For example in group theory, two groups are
\emph{the same} iff they are isomorphic; in topology, two spaces are
\emph{the same} iff they are homeomorphic; etc. In proof theory, we
have no such notion telling us when two proofs are \emph{the same},
even though Hilbert was considering this problem as a possible 24th
problem for his famous lecture~\cite{hilbert:00} in
1900~\cite{thiele:03}, before proof theory existed as a mathematical
field.

@@

This is where \emph{combinatorial proofs} come in. They have been
introduced by Hughes~\cite{hughes:pws} for classical propositional
logic as a ``syntax-free'' notion of proof, and as a potential solution
to Hilbert's 24th problem~\cite{hughes:invar} (see
also~\cite{str:hilbert:24}). The basic idea is to abstract away from
the syntax of the inference rules used in the proof and consider the
proof as a combinatorial object, more precisely as a special kind of
graph homomorphism. For example, a propositional combinatorial
proof of Peirce's law $\peirceimpliesformula$ is depicted below-left.
%
\begin{center}\vspace{.5ex}\begin{tikzpicture}\newcommand\radius{2.3}
  \begin{scope}[shift={(-\radius,0)}]{\peircecp}\end{scope}
  \begin{scope}[shift={(\radius,0)}]{\drinkerfibcoloured}\end{scope}
\end{tikzpicture}\vspace{0ex}\end{center}
%
Several authors have illustrated how syntactic proofs in various
proof systems can be translated to propositional combinatorial proofs:
for sequent proofs in~\cite{hughes:invar}, for deep inference proofs
in~\cite{str:fscd17}, for Frege systems in~\cite{str:RR-9048}, and for
tableaux systems and resolution in~\cite{acc:str:18}. This enables a
natural definition of proof identity for propositional logic: two
proofs are \emph{the same}, if they are mapped to the same
combinatorial proof.

Recently, Acclavio and Stra\ss burger extended this notion to relevant
logics~\cite{acc:str:relevant} and to modal
logics~\cite{acc:str:modal}, and Heijlties, Hughes and Stra\ss burger
have provided combinatorial proofs for intuitionistic propositional
logic~\cite{HHS:lics19}.

In this paper we advance the idea that combinatorial proofs can provide
a notion of proof identity for first order logic. \emph{First-order
combinatorial proofs} were introduced by Hughes in~\cite{hughes:fopws}.
However, even though Hughes proves soundness and completeness, the 
proof is unsatisfactory: (1) the soundness argument is long and
cumbersome, and (2) the completeness proof does not allow a syntactic
proof to be read back from a combinatorial proof. In fact, there is the
fundamental problem that not all combinatorial proofs can be obtained
as translations of sequent calculus proofs.

In this paper we solve this issue by moving to a deep inference
system. More precisely, we introduce a new proof system, called
$\FOKS$, for first-order logic, that (a) reflects every combinatorial
proof, i.e., there is a surjective mapping from proofs in $\FOKS$ to
combinatorial proofs, and (b) yields far simpler proofs of
soundness and completeness of combinatorial proofs, and (c) admits
new decomposition theorems establishing a precise correspondence
between certain syntactic inference rules and certain combinatorial
notions.

@@

The fograph of drinker formula $\drinkerformula=\veedrinkerformula$:
%
\begin{center}\vspace{2ex}\begin{tikzpicture}
\drinkerbase
\end{tikzpicture}\vspace{1ex}\end{center}
%
%
Condensed combinatorial proof of drinker formula(s):
%
\begin{center}\vspace{3ex}\begin{math}
\begin{tikzpicture}
\drinkerInline
\end{tikzpicture}
\hspace{10ex}
\begin{tikzpicture}
\veedrinkerInline
\end{tikzpicture}\end{math}\end{center}
